Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-05-10 manuscript.tex
added thermo section without dangling ends

\newcommand*{\rood}[1]{{\color{red}{#1}}}
%\newcommand*{\rood}[1]{#1}
\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
%\newcommand*{\blauw}[1]{#1}
\newcommand*{\groen}[1]{\textcolor{green}{#1}}
%\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{{\color{blue}{#1}}}
%\newcommand*{\blauwr}[1]{#1}

\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}

\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes on
%\newcommand*{\noter}[1]{}					% notes off

@@

We performed molecular dynamics simulations of four 10-base self-complementary double-stranded DNA sequences that we have previously studied by ultra-fast temperature-jump infrared spectroscopy \citep{Sanstead2016}: 5$^\prime$ATATATATAT3$^\prime$ (AT-all), 5$^\prime$GATATATATC3$^\prime$ (GC-end), 5$^\prime$ATATGCATAT3$^\prime$ (GC-core), and 5$^\prime$ATGATATCAT3$^\prime$ (GC-mix). We modeled the DNA sequences using the coarse-grained 3-Site-Per-Nucleotide v2 (3SPN.2) that uses three spherical beads to represent the phosphate, deoxyribose sugar, and nitrogenous base of each nucleotide and employs anisotropic interaction potentials to accurately treat intra-strand base-stacking, inter-strand cross-stacking, and base pairing. \citep{Hinckley2013AnHybridization} The model was parameterized against experimental data on bond lengths, bend angles, torsional angles, base step energies, and base stacking free energies, and reliably reproduces the structure, melting temperatures, persistence lengths, and sequence, salt, concentration, and temperature effects on duplex formation. \citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in very good agreement with observed experimental trends.

All calculations were performed using the LAMMPS simulation package in accordance with best practices for the 3SPN.2 model. \addref{LAMMPS} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of \rood{XX} mol/L. Solvent effects were modeled implicitly by employing Langevin dynamics \citep{Schneider1978Molecular-dynamicsTransitions} \noter{Is this the right reference for Langevin dynamics?} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s. \citep{Hinckley2013AnHybridization} \addref{A. E. Nkodo, G. M. Garnier, B. Tinland, C. Desruisseaux, L. C. Mc-Cormick, G. Drouin, and G. W. Slater,Electrophoresis22, 2424 (2001)} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius. \addref{Debye-Huckel} Simulations were performed in the NVT ensemble employing a Langevin thermostat. \addref{Langevin thermostat} Each sequence was simulated at its melting temperature -- AT-all: \rood{XX} K, GC-end: \rood{XX} K, GC-core: \rood{XX} K, GC-mix: \rood{XX} K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states. Melting temperatures for each sequence were determined empirically by \rood{XXX}. The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \addref{G. Bussi and M. Parrinello,Phys.Rev.E75, 056707 (2007).} with a 20 fs integration time step. We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state. Initial system configurations were prepared \rood{How? How specify initial configurations and apply any energy minimization?} Initial bead velocities assigned from a Maxwell-Boltzmann distribution at the system temperature. \addref{Maxwell-Boltzmann} Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores. The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.

@@

Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly} \addref{(1) Husic, B. E.; Pande, V. S. Markov State Models: From an Art to a Science. J. Am. Chem. Soc. 2018, 140, 2386?2396. (2) Pande, V. S.; Beauchamp, K.; Bowman, G. R. Everything You Wanted to Know About Markov State Models But Were Afraid to Ask. Methods 2010, 52, 99?105. (3) https://doi.org/10.33011/livecoms.1.1.5965 (4) 10.1016/j.sbi.2014.04.002} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation. MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation. Finally, a recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference. Calculations were performed using the PyEMMA software package. \addref{}

\textbf{Featurization.} Trajectories comprising the Cartesian coordinates of the DNA strands as a function of time were featurized using the MDtraj Python libraries \citep{McGibbon2015MDTraj:Trajectories} to represent the system in a manner that exposes the essential system dynamics but eliminates trivial translation and rotational invariances. We adopt pairwise distances $d(i,j)$ between the centers of mass of the 10 \rood{bases / nucleotides / what?} as a natural rototranslationally invariant featurization that represents each system configuration as the $20 \choose 2$ = 190-element vector of pairwise distances between all distinguishable bases within the complementary sequence pair. This featurization naturally accounts for intramolecular configurations via the pairwise distances between bases within the same strand and intermolecular configurations via the pairwise distances between bases on different strands. One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 190-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances. Specifically, $\left( d(i_1,j_2) = d(i_2,j_1) \right) \leftarrow 0.5\left( d(i_1,j_2) + d(i_2,j_1) \right)$ \noter{Right?}, where $i_1$ denotes the $i^\mathrm{th}$ \rood{bases / nucleotides / what?} on strand 1 and $j_2$ the $j^\mathrm{th}$ \rood{bases / nucleotides / what?} on strand 2. \citep{Sengupta2019AutomatedSelf-assembly} Finally, we elected to work with the reciprocal of the permutationally-symmetrized pairwise distances to provide higher resolution and differentiation between proximate strand configurations in the near hybridized state compared to distantly separated dissociated strands. VAMP-2 scoring -- calculation of the sum of the squared estimated eigenvalues of the transfer operator -- of trajectories under a particular featurization provides a measure of the kinetic variance carried by that featurization. \citep{Noe2013ASystems} \addref{https://doi.org/10.1021/acs.jctc.5b00553. (38) Scherer,M.K.;Husic,B.E.;Hoffmann,M.;Paul,F.;Wu,H.; Noe? F. Variational Selection of Features for Molecular Kinetics. J. Chem. Phys. 2019, 150, No. 194108. (39) Wu, H.; Noe? F. Variational Approach for Learning Markov Processes from Time Series Data. 2017} Performing VAMP-2 scoring at a lag time of $\tau$ = \rood{XX} ps and retaining the top \rood{XX} modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to \rood{XX}\% more kinetic variance than the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model.

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data. \addref{[27] Pérez-Hernández G, Paul F, Giorgino T, Fabritiis GD, Noé F. Identification of slow molecular order parameters for Mar- kov model construction. J Chem Phys. 2013; 139(1):015102. https://doi.org/10.1063/1.4811489.
[28] SchwantesCR,PandeVS.ImprovementsinMarkovStateModel Construction Reveal Many Non-Native Interactions in the Fold- ing of NTL9. J Chem Theory Comput. 2013; 9(4):2000?2009. https://doi.org/10.1021/ct300878a.
[29] Noé F, Clementi C. Kinetic Distance and Kinetic Maps from Molecular Dynamics Simulation. J Chem Theory Comput. 2015; 11(10):5002?5011. https://doi.org/10.1021/acs.jctc.5b00553.} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of \rood{0.01}, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriate short to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \addref{https://doi.org/10.1063/1.5002086} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs. We guarded against overfitting using cross-validation in which we divided the trajectory data into \rood{XX} contiguous segments of \rood{XX} ns that were then partitioned into an 80:20 training:validation split. We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single \rood{NVIDIA XXX} GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth slow mode (\rood{Fig.~S1} in the \rood{Supporting Information} \noter{What is this called in JACS? Supporting, Supplementary?}). This motivated the construction of low-dimensional embeddings of the trajectories for each of the four DNA sequences into a 5D space of leading SRV modes.

\textbf{Microstate clustering.} The 5D SRV projections of the \rood{XX} frames recorded over the course of the 1 ms MD trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering. The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was highly insensitive to the choice of the number of microstates over the range \rood{XX-YY}, motivating our selection of 200 microstate clusters for each system. \noter{Accurate statement?}

\textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM. The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda i \right|$. \addref{https://doi.org/10.33011/livecoms.1.1.5965} The implied time scale plots for the four DNA sequences are presented in \rood{Fig.~S2}. We observe convergence of the implied time scales very rapidly with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns. At this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively. The identification of $(i-1)$ implied time scales implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates. We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix. \addref{(43) Röblitz, S.; Weber, M. Fuzzy Spectral Clustering by PCCA+: Application to Markov State Models and Data Classification. Adv. Data Anal. Classif. 2013, 7, 147?179.
(44) Deuflhard, P.; Weber, M. Robust Perron Cluster Analysis in Conformation Dynamics. Linear Algebra Appl. 2005, 398, 161?184.
(45) Kube, S.; Weber, M. A Coarse Graining Method for the Identification of Transition Rates Between Molecular Conformations. J. Chem. Phys. 2007, 126, No. 024103.} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mid systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates. These macrostate MSMs constitute our terminal kinetic models. We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \addref{https://doi.org/http://dx.doi.org/10.1063/1.3565032., https://doi.org/10.1073/pnas.0905466106., https://doi.org/10.33011/livecoms.1.1.5965}. This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$. We present in \rood{Fig.~S3} the CK tests for each DNA sequence to demonstrate that the $\tau$ = 1.2 ns models performs excellently in predicting transition probabilities out to $k \tau$ = \rood{XX} ns, validating the Markovain nature and kinetic validity of the four kinetic models.

@@

Each DNA oligonucleotide was purchased from Integrated DNA Technologies (IDT) at desalt grade purity. Oligonucleotides were purified with 3 kD cutoff centrifugal filters (Amicon). All labile protons were exchanged in deuterium oxide (\ce{D2O}, Cambridge Isotopes, 99.9\%). Oligonucleotides were prepared at a total strand concentration of 2 mM in 50 mM pD 7.2 sodium phosphate buffer with 240 mM NaCl and 18 mM \ce{MgCl2}.

@@

The details of the technique and processing used to acquire temperature-jump infrared (T-jump IR) data have been described previously.~\addref{1.	Chung, H. S.; Khalil, M.; Smith, A. W.; Tokmakoff, A., Transient two-dimensional IR spectrometer for probing nanosecond temperature-jump kinetics. Rev. Sci. Instrum. 2007, 78, 063101.
2.	Jones, K. C.; Ganim, Z.; Tokmakoff, A., Heterodyne-detected dispersed vibrational echo spectroscopy. J. Phys. Chem. A 2009, 113, 14060-14066.
3.	Jones, K. C.; Ganim, Z.; Peng, C. S.; Tokmakoff, A., Transient two-dimensional spectroscopy with linear absorption corrections applied to temperature-jump two-dimensional infrared. JOSA B 2012, 29, 118-129.
} Briefly, heating was initiated through optical excitation of the O-D stretch overtone transition of \ce{D2O}. The 1.98 $\mu$m pulses (5 ns, 20 mJ, 20 Hz) used for heating were generated from the frequency-doubled output of a Nd:YAG (YG 980, Quantel) sent through an optical parametric oscillator (OPO). Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump. T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\addref{2.	Jones, K. C.; Ganim, Z.; Tokmakoff, A., Heterodyne-detected dispersed vibrational echo spectroscopy. J. Phys. Chem. A 2009, 113, 14060-14066.} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between -10 and 10 fs. t-HDVE spectra were acquired with a \rood{ZZZZ} \noter{Is this correct?} polarization scheme. Data are presented using the recovered T-jump dispersed pump-probe (t-DPP) spectrum. t-DPP data are reported as the difference spectra with the 49 or 50 ms spectrum, which corresponds the initial temperature spectrum.

@@

\begin{figure}[ht!]
	\begin{center} 
        %\includegraphics[width=150mm, scale=1]{Fig6.pdf}
        \caption*{TOC image}
        \label{fig:TOC}
	\end{center}
\end{figure}

>>>>

\newcommand*{\rood}[1]{{\color{red}{#1}}}
\newcommand*{\bood}[1]{{\color{blue}{#1}}}
\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\groen}[1]{\textcolor{green}{#1}}

\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes 
%\newcommand*{\noter}[1]{}					% notes off

@@

We performed molecular dynamics simulations of four 10-base self-complementary double-stranded DNA sequences that we have previously studied by ultra-fast temperature-jump infrared spectroscopy \citep{Sanstead2016}: 5$^\prime$-ATATATATAT-3$^\prime$ (AT-all), 5$^\prime$-GATATATATC3-3$^\prime$ (GC-end), 5$^\prime$-ATATGCATAT-3$^\prime$ (GC-core), and  5$^\prime$-ATGATATCAT-3$^\prime$ (GC-mix). We modeled the DNA sequences using the coarse-grained 3-Site-Per-Nucleotide v2 (3SPN.2) that uses three spherical beads to represent the phosphate, deoxyribose sugar, and nitrogenous base of each nucleotide and employs anisotropic interaction potentials to accurately treat intra-strand base-stacking, inter-strand cross-stacking, and base pairing. \citep{Hinckley2013AnHybridization} The model was parameterized against experimental data on bond lengths, bend angles, torsional angles, base step energies, and base stacking free energies, and reliably reproduces the structure, melting temperatures, persistence lengths, and sequence, salt, concentration, and temperature effects on duplex formation. \citep{Hinckley2013AnHybridization} The model enables access to millisecond time scales and has been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping.\citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping} Although the 3SPN.2 model was not directly parameterized against dynamical experimental data, we will show below that the predicted sequence-dependent kinetics and relaxation are, within a corrective scaling factor, in very good agreement with observed experimental trends.

All calculations were performed using the LAMMPS simulation package (\url{http://lammps.sandia.gov}) in accordance with best practices for the 3SPN.2 model. \citep{Plimpton1997Short-RangeDynamics} A single pair of self-complementary sequences were placed in a cubic periodic box with side length 7.8 nm corresponding to a single-strand concentration of 7 mol/L. Solvent effects were modeled implicitly by employing Langevin dynamics \citep{DunwegB.Paul1991BrownianNumbers.} with an experimentally motivated per-site friction coefficient of $\xi$ = 9.94 $\times$ 10$^{-11}$ m$^2$/s. \citep{Nkodo2001DiffusionElectrophoresis, Hinckley2013AnHybridization} We specified a 240 mM implicit salt concentration and treated electrostatic interactions using the Debye-H{\"u}ckel with a 5 nm cutoff radius. \citep{Debye1923ZurElektrolyte} Simulations were performed in the NVT ensemble employing a Langevin thermostat. \citep{Schneider1978Molecular-dynamicsTransitions}. Each sequence was simulated at its melting temperature -- AT-all: 309 K, GC-end: 317 K, GC-core: 324 K, GC-mix: 324 K -- in order to maximize the number of spontaneous transitions between dissociated and hybridized states. Melting temperatures for each sequence were determined empirically by comparing the ratio of hybridized to dissociated populations over a 10 K temperature ramp centered on the Nearest Neighbors model predicted melting temperature. The temperature at which the hybridized and dissociated populations were closest was selected. The Langevin equations of motion were integrated using the scheme of Bussi and Parrinello \citep{Bussi2007AccurateDynamics} with a 20 fs integration time step. We performed 40 independent simulations for each of the four sequences with half of the runs initialized from the hybridized state and half from the dissociated state. The initial hybridized state was set, according to 3SPN.2 documentation, to the coordinates of Arnott et al.\ ~\citep{Arnott1976CRCBiology} The dissociated state was generated by shifting one oligomer from the hybridized state by 1 nm in the x, y, and z directions. The magnitude of this shift did not affect the results so long as all native WC bonds were sufficiently separated. \noteb{No energy minimization (no 3spn2 precedent on this). The lammps default emin techniques may not work for this type of CG forcefield. Given the sampling it seems that  1 $\mu$s of equilibration should be sufficient?} Initial bead velocities were assigned from a Maxwell-Boltzmann distribution at the system temperature. \addref{Maxwell-Boltzmann -- \noteb{not sure what to use here, usually not cited?}} Each simulation was conducted for 26 $\mu$s and frame saved to disc every 100 ps, requiring $\sim$24 CPU-hours on 28$\times$Intel E5-2680v4 CPU cores. The first 1 $\mu$s of each run was discarded for equilibration providing us with 40$\times$25 $\mu$s = 1 ms of simulation data for each sequence, during which time we observed 55-100 hybridization/dehybridization events.

@@

Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly} \citep{ Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionV1.0} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation. MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation. Finally, a recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference. Calculations were performed using the PyEMMA software package. \citep{Scherer2015PyEMMAModels}

\textbf{Featurization.} Trajectories comprising the Cartesian coordinates of the DNA strands as a function of time were featurized using the MDtraj Python libraries \citep{McGibbon2015MDTraj:Trajectories} to represent the system in a manner that exposes the essential system dynamics but eliminates trivial translation and rotational invariances. We adopt pairwise distances $d(i,j)$ between the centers of mass of the 10 bases as a natural rototranslationally invariant featurization that represents each system configuration as the $(10 \times 10)$ 100-element vector of pairwise distances between all distinguishable bases within the complementary sequence pair. This featurization naturally accounts for intermolecular configurations via the pairwise distances between bases on different strands. \noteb{previous text had suggested we used the 190-dim features, but we actually used $100-45 = 55$-dim features since the intramolecular features contributed no additional kinetic variance (I've updated accordingly)} One additional symmetry arises from the self-complementary nature of these sequences -- the sense and anti-sense strands in each pair are identical -- such that the representation of the system under our featurization should remain unchanged upon inverting the arbitrary labeling of strand ``1'' and strand ``2''.\citep{Sengupta2019AutomatedSelf-assembly} The 100-element pairwise distance vector is not invariant to this permutation, but can easily be made so via a simple symmetrization operation in which each of the $10 \choose 2$ =  45 intermolecular pairwise distances are replaced by the mean of the two permutationally invariant distances. Specifically, $\left( d(i_1,j_2) = d(i_2,j_1) \right) \leftarrow 0.5\left( d(i_1,j_2) + d(i_2,j_1) \right)$, where $i_1$ denotes the $i^\mathrm{th}$ base on strand 1 and $j_2$ the $j^\mathrm{th}$ base on strand 2. \citep{Sengupta2019AutomatedSelf-assembly} Finally, we elected to work with the reciprocal of the permutationally-symmetrized pairwise distances to provide higher resolution and differentiation between proximate strand configurations in the near hybridized state compared to distantly separated dissociated strands. VAMP-2 scoring -- calculation of the sum of the squared estimated eigenvalues of the transfer operator -- of trajectories under a particular featurization provides a measure of the kinetic variance carried by that featurization. \citep{Noe2013ASystems, Noe2015KineticSimulation, Scherer2019VariationalKinetics, Wu2020VariationalData} Performing VAMP-2 scoring at a lag time of $\tau$ = 1.2 ns and retaining the top 5 modes, reveals that the reciprocal permutationally-symmetrized pairwise distances can carry up to twice the kinetic variance as the non-reciprocal distances, suggesting that the higher resolution offered at close intermolecular distances can indeed boost the dynamical representational power of the model.

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriate short to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs. We guarded against overfitting using 5-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits. We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA GK210GL GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively \noteb{not sure the best way to explain this succinctly}. (Figure 1 in the Supporting Information). This motivated the construction of a n-dimensional embedding corresponding to n leading SRV modes retained for each sequence.

\textbf{Microstate clustering.} The SRV projections of the 10 million frames recorded over the course of the 1 ms MD trajectories collected for each DNA sequence were then clustered into microstates using k-means clustering. The VAMP-2 score of the microstate transition matrix constructed for each sequence at the selected $\tau$ = 1.2 ns was highly insensitive to the choice of the number of microstates over the range 100-1000, motivating our selection of 200 microstate clusters for each system.

\textbf{Macrostate clustering.} The 200 microstates comprising each system were finally coarsened into our terminal macrostate MSM. The microstate transition matrix for each system was computed at a range of lag times $\tau$ and then diagonalized to recover the corresponding eigenvalues $\lambda_i$ and associated implied time scales $t_i = -\tau / \ln \left| \lambda i \right|$.~\citep{Wehmeyer2019IntroductionV1.0} The implied time scale plots for the four DNA sequences are presented in SI Figure 2. We observe convergence of the implied time scales very rapidly with lag time $\tau$ for all systems, motivating the construction of high resolution macrostate MSMs at a lag time $\tau$ = 1.2 ns. At this choice of lag time, we recover 5, 4, 2, and 1 implied time scales for the AT-all, GC-end, GC-core, and GC-mid systems, respectively. The identification of $(i-1)$ implied time scales \noteb{(may be 1-2 additional converging, but these correspond to diffusive processes that are irrelevant or unhelpful in determining the hybridization mechanisms. Still, this might come up when looking at the the SI Fig 2} implies the presence of $(i-1)$ slow modes and motivates the coarsening of the system into $i$ macrostates. We estimate these $i$ macrostates by applying PCCA+ spectral clustering to the leading $(i-1)$ eigenvectors of the microstate transition matrix.~\citep{Roblitz2013FuzzyClassification, Weber2018ImplicationsSimulation, Kube2007AConformations} We then estimate the corresponding 6, 5, 3, and 2 macrostate transition matrices $\mathbf{P}$ for the AT-all, GC-end, GC-core, and GC-mid systems, respectively, by projecting the MD simulation trajectories into these discrete macrostates. These macrostate MSMs constitute our terminal kinetic models. We validate the Markovian nature of the four MSMs by subjecting them to the Chapman-Kolmogorov (CK) test \citep{Noe2009ConstructingSimulations,Phys2011MarkovValidation,  Wehmeyer2019IntroductionV1.0}. This test asserts that the transition matrix for a Markovian (i.e., memoryless) MSM constructed at a lag time $\tau$ should satisfy the condition $\mathbf{P}(k \tau) = \mathbf{P}^k(\tau)$, which states that $k$ successive applications of the transition matrix constructed at a lag time $\tau$should be equivalent to a single application of the transition matrix constructed at a lag time $k \tau$. We present in SI Figure 3 the CK tests for each DNA sequence to demonstrate that the $\tau$ = 1.2 ns models performs excellently in predicting transition probabilities out to $k \tau$ = 7.2 ns, validating the Markovian nature and kinetic validity of the four kinetic models.

@@

Each DNA oligonucleotide was purchased from Integrated DNA Technologies (IDT) at desalt grade purity. Oligonucleotides were purified with 3 kD cutoff centrifugal filters (Amicon). All labile protons were exchanged in deuterium oxide (\ce{D2O}, Cambridge Isotopes, 99.9\%). Oligonucleotides were prepared at a total strand concentration of 2 mM in 50 mM pD 7.2 sodium phosphate buffer with 240 mM NaCl and 18 mM \ce{MgCl2}. Prior to each measurement, DNA solutions were placed in a water bath at 90$^{\circ}$ C for 3 minutes and then cooled to room temperature under ambient conditions.

@@

The details of the technique and processing used to acquire temperature-jump infrared (T-jump IR) data have been described previously.~\citep{Chung2007TransientKinetics,Jones2009Heterodyne-detectedSpectroscopy,Jones2012TransientInfrared} Briefly, heating was initiated through optical excitation of the O-D stretch overtone transition of \ce{D2O}. The 1.98 $\mu$m pulses (5 ns, 20 mJ, 20 Hz) used for heating were generated from the frequency-doubled output of a Nd:YAG (YG 980, Quantel) later sent through an optical parametric oscillator (OPO). Nonlinear IR spectra are collected from 5 ns to 50 ms delays after the T-jump. T-jump heterodyne-detected vibrational echo (t-HDVE) IR spectra were acquired with Fourier transform spectral interferometry,~\citep{Jones2009Heterodyne-detectedSpectroscopy} where the delay between the local oscillator (LO) and DVE signal was scanned in 5 fs steps between -10 and 10 fs. t-HDVE spectra were acquired with a parallel pulse polarization scheme. Data are presented using the recovered T-jump dispersed pump-probe (t-DPP) spectrum. t-DPP data are reported as the difference spectra with the 49 or 50 ms spectrum, which corresponds the initial temperature spectrum.

@@

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=150mm, scale=1]{TOC.pdf}
        \caption*{TOC image \noteb{could consider adding an experimental aspect to this?}}
        \label{fig:TOC}
	\end{center}
\end{figure}
